---
title: 'Project #1 - Data Exploration'
author: "Khalid Abdulshafi"
date: "1/30/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(scales)
```

```{r}
pro <- read.csv("ProjectA_Listings2013.csv")
```

```{r}
#Check is NA values in loan_status
any(is.na(pro$loan_status))

```

```{r}
str(pro)
summary(pro)
```

```{r}
#factor categorical variables
pro$loan_status <- as.factor(pro$loan_status)
pro$loan_origination_date <- as.factor(pro$loan_origination_date)
pro$prosper_rating <- as.factor(pro$prosper_rating)
pro$listing_term <- as.factor(pro$listing_term)
pro$scorex <- as.factor(pro$scorex)
pro$prosper_score <- as.factor(pro$prosper_score)
pro$income_range <- as.factor(pro$income_range)
pro$income_verifiable <- as.factor(pro$income_verifiable)
pro$employment_status_description <- as.factor(pro$employment_status_description)
pro$occupation <- as.factor(pro$occupation)
pro$borrower_state <- as.factor(pro$borrower_state)
pro$borrower_city <- as.factor(pro$borrower_city)
pro$lender_indicator <- as.factor(pro$lender_indicator)
pro$is_homeowner <- as.factor(pro$is_homeowner)
```

```{r}
#delete columns
pro$loan_status_description <- NULL
pro$income_range_description <- NULL
pro$listing_category_id <- NULL
```

```{r}
#We want to take a closer look at borrowers, not lenders
borrower_subset <- subset(pro, pro$lender_indicator == 0)
str(borrower_subset)
summary(borrower_subset)
```

```{r}
mean_num_yrs <- round(mean(borrower_subset$number_of_days)/365)

rel_frq_loan_status <- table(borrower_subset$loan_status)/length(borrower_subset$loan_status)

mean_num_days_vs_list_term <- tapply(borrower_subset$number_of_days, borrower_subset$listing_term, mean, na.rm = TRUE)
```

## Key Takeaways

-   It takes an average of `r round(mean_num_days_vs_list_term[1]/365)` years to pay off a 1-year loan, `r round(mean_num_days_vs_list_term[2]/365)` for a 3-year loan, and `r round(mean_num_days_vs_list_term[3]/365)` for 5-year loan.
-   `r percent(rel_frq_loan_status[1])` of the loans are currently active, `r percent(rel_frq_loan_status[2])` have been charged off, `r percent(rel_frq_loan_status[3])` have been defaulted on, and `r percent(rel_frq_loan_status[4])` have been paid off. 
-   The average loan extended to borrowers is `r round(mean(borrower_subset$amount_funded, scientific = FALSE))`.
- The average personal loan interest rate is `r round(mean(borrower_subset$borrower_rate))`%.
-   Debtors come from 48 states and 6531 cities.

## Grouped Analysis
```{r}
# is_homeowner vs laon status
# income verifiable vs loan status
# employment status vs laon status
# income range vs loan status
# prosper rating vs loan status
# scorex vs loan status
# borrower state vs loan status
```

## Including Plots
```{r}

```

## Questions
-   Why do borrowers take an average of 3 years to pay off a 1-year loan?